---
title: "Correct errors in R code"
author: "Copilot + Svetlana"
date: "2025-07-14"
output: html_document
css: "copilot.css"
---

```{r setup, echo=FALSE}
# let rmarkdown not stop on error and knit the document
knitr::opts_chunk$set(error=TRUE)
```

## R errors

I was too lazy learning R and asked Copilot to do the R course homework for me. Unfortunately, the code contained mistakes, and I got errors and unexpected results. (Always check what your LLM gives you.)

R is not always very explicit in its error messages, and sometimes it even passes an error silently on.

Let's try to understand the errors and fix that faulty code! Feel free to work in pairs because 2 brains is better than one :-) :-)

### 1. Variables naming

```{r}
4u <- "There is a dataset for you"
```

```{r}
lucky-number <- 42
```

```{r}
data <- 1000
print(Data)
```

```{r}
print <- "Printing answer.."
help(print)
```

### 2. Basic data types and operations

```{r}
pi <- as.numeric("3.14p")
pi
```

```{r}
n <- 10
print("Reporting " + n + " top results")
```

```{r}
x <- list(1, 2, 3)
sum(x)
```

```{r}
f <- factor(c("10", "20", "30"))
as.numeric(f)
```

```{r}
colors <- factor(c("red", "blue", "green"))
colors[4] <- "yellow"
print(colors)
```

### 3. Scope

```{r}
func = function(){
  age = 18    
} 

print(age)
```

```{r}
# we want to double the i every time (expect 32)
for(i in 1:5){
  value <- i * 2
}
print(value)
```

### 4. Vector Operations

```{r}
# we want to subset x based on mask and accidentally chose the wrong one - no error
x <- c(5, 10, 15, 20, 25)
mask <- c(TRUE, FALSE)
result <- x[mask]
print(result)
```

### 5. Indexing

```{r}
x <- 1:4
x[0]
```

```{r}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35)
)

# take mean age of persons
mean(df[2])
```

```{r}
# filter for those who are called Alice
print(df[name == "Alice",])
```

### 6. Missing values

```{r}
# imagine the real data has thousands of points and we cannot check missing value by eye
data <- c(1, 2, NA, 4, 5)
mean(data)
```

```{r}
# here we are trying to deterimne if data contains NA
data <- c(1, 2, NA, 4, 5)
if(data == NA) {
  print("Found NA")
}
```

### 7. Namespace

These two chunks of code should give the same result, or?

```{r}
x <- 1:100
filter(x, rep(1, 3))
```

```{r}
library(tidyverse)
x <- 1:100
filter(x, rep(1, 3))
```
