---
title: "Correct errors in R code"
author: "Copilot + Svetlana"
date: "2025-07-14"
output: html_document
---

```{r setup, echo=FALSE}
# let rmarkdown not stop on error and knit the document
knitr::opts_chunk$set(error=TRUE)
```

## R errors

I was too lazy learning R and asked Copilot to do the R course homework for me. Unfortunately, the code contained mistakes, and I got errors and unexpected results. (Always check what your LLM gives you.)

R is not always very explicit in its error messages, and sometimes it even passes an error silently on.

Let's try to understand the errors and fix that faulty code! Feel free to work in pairs because 2 brains is better than one :-) :-)

### 1. Variables naming

Variable cannot start with a number (but can end with one):

```{r}
#2bit <- "hg38.2bit"
bit2 <- "hg38.2bit"
```

Variable cannot contain `-`(but can contain `_`):

```{r}
#two-bit-file <- "hg38.2bit"
two_bit_file <- "hg38.2bit"
```

Case sensitive typos:

```{r}
twoBit <- "hg38.2bit"
#print(TwoBit)
print(twoBit)
```

Do not overwrite built-in names:

```{r}
#print <- "Printing answer.."
# if you ran the line above, make it good again with:
#rm(print)
print_statement <- "Printing answer.."
help(print)
```

### 2. Basic data types and operations

Cannot coerce letters to numeric:

```{r}
#pi <- as.numeric("3.14p")
pi <- as.numeric("3.14")
pi
```

Cannot use "+" with strings (unlike python). Use `paste()`.

```{r}
n <- 10
#print("Reporting " + n + " top results")
paste("Reporting ", n , " top results")
```

To convert vector to list, use `as.list()`. `list()` will give you a list oflength 1 with your vector as the first element.

```{r}
#my_list_1_to_5 <- list(1:5)
my_list_1_to_5 <- as.list(1:5)
my_list_1_to_5[4]
```

Vectorized functions do not work on a list. 2 ways:

1.  Convert list to a vector with `unlist()`
2.  Use function `do.call()` on the list.

Of course, it only works if the list can be coerced to numeric (does not contain characters).

```{r}
x <- list(1, 2, 3)
#sum(x)
sum(unlist(x))
# or
do.call(sum,x)
```

Factors are enumerated categories and thus are of class numeric and start with 1. Levels of the factor are of class character.

```{r}
f <- factor(c("10", "20", "30"))
#as.numeric(f)
as.numeric(levels(f))
```

To add a new category to a factor, make sure that it is present in levels first.

```{r}
mycolors <- factor(c("red", "blue", "green"))
# first change the levels
levels(mycolors) <- c(levels(mycolors), "yellow")
mycolors[4] <- "yellow"
print(mycolors)
```

### 3. Scope

Variables assigned inside functions are not exported to global environment.

```{r}
#func = function(){
#  age = 18    
#} 
#print(age)

func = function(){
  age = 18    
  print(age)
} 

func()
```

We have 2 variables: a counts `i` which is only used to count the rounds, and the `value` which we want to modify.

It helps to introduce a print statement to see which meaning `i` and `value` have in each cycle.

Also, make sure to reset the value in the beginning - otherwise it will keep multiplying every time you run this cell.

```{r}
#reset the value
value <- 1

# we want to double the i every time (expect 32)
for(i in 1:5){
  print(paste("i:",i))
  print(value)
  value <- value * 2
}
print("final value:")
print(value)
```

### 4. Vector Operations

Remember that vectors are recycled and it can lead to unintended result.

```{r}
# we want to subset x based on mask and accidentally chose the wrong one - no error
x <- c("good", "bad", "good", "good", "bad")
#mask <- c(TRUE, FALSE)
mask <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
result <- x[mask]
print(result)
```

### 5. Indexing

Remember that R, unlike most languages (but like humans) starts counting at 1, not at 0.

```{r}
x <- 1:4
#x[0]
x[1]
```

In case of data.frames, remember that single brackets return a `data.frame` (like in case of lists, they return a list). To get actual vector of "age", to which we can apply vectorized functions such as `sum()`, use double brackets.

```{r}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35)
)

# take mean age of persons
#mean(df[2])
mean(df[[2]])
```

In base R, we need to tell R, to which data frame `name` belongs.

```{r}
# filter for those who are called Alice
#print(df[name == "Alice",])
print(df[df$name == "Alice",])
```

### 6. Missing values

It is often safer to include na.rm=TRUE.

```{r}
# imagine the real data has thousands of points and we cannot check missing value by eye
data <- c(1, 2, NA, 4, 5)
#mean(data)
mean(data, na.rm = TRUE)
```

Potentially any missing value is different, so you cannot compare NA to NA. Thus we usually need a dedicated function `is.na()` which checks for missing value. This function returns a vector of `TRUE` and `FALSE`, indicating which elements in a vector are missing.

For this example, however, we only want to know if the vector contains `NA` at all. Remeber the `%in%` operator?

```{r}
# here we are trying to determine if data contains NA
data <- c(1, 2, NA, 4, 5)
#if(data == NA) {
  #print("Found NA")
#}
if(NA %in% data) {
  print("Found NA")
}
```

### 7. Namespace

These two chunks of code should give the same result, or?

The answer is given to you by modern R already: at the moment you load the `tidyverse` library, it will warn you that `filter()` function is masked. It is safer to use `package::function()` syntax, especially with functions which names are common.

```{r}
x <- 1:100
#filter(x, rep(1, 3))
stats::filter(x, rep(1, 3))
```

```{r}
library(tidyverse)
x <- 1:100
#filter(x, rep(1, 3))
stats::filter(x, rep(1, 3))
```
