---
title: "R Basics - Pre-course Homework"
author: "CRC1678 INF"
date: "2025-07-14"
output: html_document
css: "copilot.css"
---


Welcome to R! This exercise will help you understand basic R concepts before our class. Run each code chunk and try to understand the output. Feel free to experiment by adding more chunks.

---

### Some useful RStudio Keyboard Shortcuts

Large RStudio cheatsheet can be found at [Posit](https://posit.co/wp-content/uploads/2022/10/rstudio-ide-1.pdf).

**General**

| Action | Windows/Linux | Mac |
|--------|--------------|-----|
| Show all shortcuts | `Alt + Shift + K` |  `Option + Shift + K` |

**Console (lower left window by default)**

| Action | Windows/Linux | Mac |
|--------|--------------|-----|
| Autocompletion | `Tab` | `Tab` |
| Move cursor to console | `Ctrl + 2` | `Cmd + 2` |
| Search command history | `Ctrl + arrow-up` | `Cmd + arrow-up` |
| Interrupt current command | `Esc` | `Esc` |


**Source (upper left window by default)**

| Action | Windows/Linux | Mac |
|--------|--------------|-----|
| Move cursor to source | `Ctrl + 1` | `Cmd + 1` |
| Duplicate current line up or down | `Alt + Shift + Up/Down` | `Option + Shift + Up/Down` |
| Delete the current line | `Ctrl + D` | `Cmd + D` |
| Comment/Uncomment lines | `Ctrl + Shift + C` | `Cmd + Shift + C` |
| Run current line/selection | `Ctrl + Enter` | `Cmd + Enter` |
| Insert R chunk | `Ctrl + Alt + I` | `Cmd + Option + I` |
| Run current chunk | `Ctrl + Shift + Enter` | `Cmd + Shift + Enter` |
| Save file | `Ctrl + S` | `Cmd + S` |
| Knit document | `Ctrl + Shift + K` | `Cmd + Shift + K` |

---


### 1. Variable Assignment

In R, we can store values in variables. The preferred way is using `<-`, though `=` also works.

```{r}
# Try these assignments
x <- 5
y = 10

# Print the values
print(x)
print(y)

# Calculate with variables
z <- x + y
print(z)
```

Your turn - create some variables with different numbers:
```{r}
# Write your code here

```

## 2. Basic Data Types

R has several basic data types:

```{r}
# Numeric (numbers with decimals)
height <- 1.75
print(class(height))

# Integer (whole numbers)
age <- 25L  # L makes it explicitly integer
print(class(age))

# Character (text)
name <- "Alice"
print(class(name))

# Logical (TRUE/FALSE)
is_student <- TRUE
print(class(is_student))
```

Try creating variables of different types:
```{r}
# Write your code here

```

## 3. Vectors

Vectors are collections of elements of the same type:

```{r}
# Create numeric vector
numbers <- c(1, 2, 3, 4, 5)
print(numbers)

# Create character vector
fruits <- c("apple", "banana", "orange")
print(fruits)

# Vector operations
numbers + 2  # adds 2 to each element
numbers * 3  # multiplies each element by 3

# Access elements
print(numbers[1])  # First element
print(fruits[2])   # Second element
```

Create your own vectors and try operations:
```{r}
# Write your code here

```

## 4. Basic Statistics

R has built-in functions for basic statistics:

```{r}
data <- c(15, 23, 42, 31, 58)

mean(data)    # Average
median(data)  # Middle value
sum(data)     # Sum of all values
length(data)  # Number of elements
max(data)     # Maximum value
min(data)     # Minimum value
```

Try these functions with your own numbers:
```{r}
# Write your code here

```

## 5. Data Frames

Data frames are like spreadsheets in R:

```{r}
# Create a simple data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 21),
  grade = c(85, 92, 78)
)

# Look at the data
print(students)

# Access columns
print(students$name)
print(students[, "age"])

# Access rows
print(students[1, ])  # First row
```

Create your own data frame about any topic:
```{r}
# Write your code here

```

## 6. Missing Values

R uses `NA` to represent missing values:

```{r}
# Vector with missing values
scores <- c(85, NA, 92, 78, NA)

# Check for missing values
is.na(scores)

# Calculate mean with missing values
mean(scores)  # Returns NA
mean(scores, na.rm = TRUE)  # Ignores NA
```

Try working with missing values:
```{r}
# Write your code here

```

## 7. Basic Plotting

Let's create a simple plot:

```{r}
# Create some data
x <- 1:10
y <- x^2

# Make a basic plot
plot(x, y, 
     main = "Simple Plot",
     xlab = "X axis",
     ylab = "Y axis")
```

Try creating your own plot:
```{r}
# Write your code here

```

## Exercises

1. Create a vector of 5 numbers and calculate their mean and sum.
2. Create a data frame containing information about 3 movies (title, year, rating).
3. Create a simple plot using different numbers than shown above.
4. Try adding two vectors of the same length.
5. Create a vector with some missing values and calculate its mean.

Submit your completed homework before the class. Remember: experimentation is key to learning R!
```

This Rmd file:
- Introduces key concepts progressively
- Includes examples and explanations
- Provides practice spaces
- Has concrete exercises
- Encourages experimentation

Would you like me to modify any sections or add more concepts?// filepath: pre_course_basics.Rmd
---
title: "R Basics - Pre-course Homework"
author: "Your Name"
date: "2025-06-13"
output: html_document
---

## Introduction

Welcome to R programming! This homework will help you understand basic R concepts before our class. Run each code chunk and try to understand the output. Add your own experiments in the empty chunks provided.

## 1. Variable Assignment

In R, we can store values in variables. The preferred way is using `<-`, though `=` also works.

```{r}
# Try these assignments
x <- 5
y = 10

# Print the values
print(x)
print(y)

# Calculate with variables
z <- x + y
print(z)
```

Your turn - create some variables with different numbers:
```{r}
# Write your code here

```

## 2. Basic Data Types

R has several basic data types:

```{r}
# Numeric (numbers with decimals)
height <- 1.75
print(class(height))

# Integer (whole numbers)
age <- 25L  # L makes it explicitly integer
print(class(age))

# Character (text)
name <- "Alice"
print(class(name))

# Logical (TRUE/FALSE)
is_student <- TRUE
print(class(is_student))
```

Try creating variables of different types:
```{r}
# Write your code here

```

## 3. Vectors

Vectors are collections of elements of the same type:

```{r}
# Create numeric vector
numbers <- c(1, 2, 3, 4, 5)
print(numbers)

# Create character vector
fruits <- c("apple", "banana", "orange")
print(fruits)

# Vector operations
numbers + 2  # adds 2 to each element
numbers * 3  # multiplies each element by 3

# Access elements
print(numbers[1])  # First element
print(fruits[2])   # Second element
```

Create your own vectors and try operations:
```{r}
# Write your code here

```

## 4. Basic Statistics

R has built-in functions for basic statistics:

```{r}
data <- c(15, 23, 42, 31, 58)

mean(data)    # Average
median(data)  # Middle value
sum(data)     # Sum of all values
length(data)  # Number of elements
max(data)     # Maximum value
min(data)     # Minimum value
```

Try these functions with your own numbers:
```{r}
# Write your code here

```

## 5. Data Frames

Data frames are like spreadsheets in R:

```{r}
# Create a simple data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 21),
  grade = c(85, 92, 78)
)

# Look at the data
print(students)

# Access columns
print(students$name)
print(students[, "age"])

# Access rows
print(students[1, ])  # First row
```

Create your own data frame about any topic:
```{r}
# Write your code here

```

## 6. Missing Values

R uses `NA` to represent missing values:

```{r}
# Vector with missing values
scores <- c(85, NA, 92, 78, NA)

# Check for missing values
is.na(scores)

# Calculate mean with missing values
mean(scores)  # Returns NA
mean(scores, na.rm = TRUE)  # Ignores NA
```

Try working with missing values:
```{r}
# Write your code here

```

## 7. Basic Plotting

Let's create a simple plot:

```{r}
# Create some data
x <- 1:10
y <- x^2

# Make a basic plot
plot(x, y, 
     main = "Simple Plot",
     xlab = "X axis",
     ylab = "Y axis")
```

Try creating your own plot:
```{r}
# Write your code here

```

## Exercises

1. Create a vector of 5 numbers and calculate their mean and sum.
2. Create a data frame containing information about 3 movies (title, year, rating).
3. Create a simple plot using different numbers than shown above.
4. Try adding two vectors of the same length.
5. Create a vector with some missing values and calculate its mean.

Submit your completed homework before the class. Remember: experimentation is key to learning R!
