---
title: "Correct errors in R code"
author: "Copilot + Svetlana"
date: "2025-07-14"
output: html_document
css: "copilot.css"
---

## R fails

I was too lazy learning R and asked Copilot to do the R course homework for me. Unfortunately, the code contained mistakes, and I got strage errors trying to run it. (Always check what your LLM gives you.)

R is not always very explicit in its error messages, and sometimes it even passes an error silently on.

Let's try to understand the errors and fix that faulty code! We will work in pairs because 2 brains is better than one :-) :-)

### 1. Variables assignment

```{r}
4u <- "There is a dataset for you"
```

```{r}
lucky-number <- 42
```

```{r}
data <- 1000
print(Data)
```

```{r}
print <- "Printing answer.."
help(print)
```

### 1. Variable Assignment and Basic Operations

Problem: What's wrong with this calculation?

```{r}
x <- 5
y = 3
z < - x + y
print(z)
```

```{r}
as.numeric("3.14p")
```

### scope

<https://www.geeksforgeeks.org/r-language/scope-of-variable-in-r/>

```{r}
func = function(){
  age = 18    
} 

print(age)
```

### 2. Vector Operations

Problem: Why doesn't this vector operation work as expected?

```{r}
numbers <- c(1, 2, 3, 4, 5)
text <- c("a", "b", "c")
result <- numbers + text
print(result)
```

### 3. Data Frame Indexing

Problem: Fix the indexing error:

```{r}
df <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35)
)
print(df[name == "Alice",])
```

### 4. Function Arguments

Problem: Why does this function fail?

```{r}
calculate_mean <- function(x, na.rm) {
  mean(x)
}
data <- c(1, 2, NA, 4, 5)
calculate_mean(data)
```

### 5. Package Usage

Problem: Identify the package-related issue:

```{r}
data <- ggplot(mtcars, aes(x = mpg, y = wt)) +
  geom_point()
```

### 6. Loop Variable Scope

Problem: Why doesn't this loop work as expected?

```{r}
total = 0
for(i in 1:5){
  value <- i * 2
}
print(value)
```

### 7. Data Type Conversion

Problem: Fix the type conversion issue:

```{r}
numbers <- c("1", "2", "3", "4", "5")
mean(numbers)
```

### 8. Missing Values

Problem: What's wrong with this NA handling?

```{r}
data <- c(1, 2, NA, 4, 5)
if(data == NA) {
  print("Found NA")
}
```

### 9. Factor Levels

Problem: Why can't we add new categories?

```{r}
colors <- factor(c("red", "blue", "green"))
colors[4] <- "yellow"
print(colors)
```

### 10. String Manipulation

Problem: Fix the string concatenation:

```{r}
first_name <- "John"
last_name <- "Doe"
full_name <- first_name + last_name
print(full_name)
```

## Solutions

<!-- Add solutions section that you can reveal during class discussion -->

<!-- Each problem has specific learning points: -->

1.  Correct operator for assignment is \<- (no spaces)
2.  Can't perform arithmetic operations between numeric and character vectors
3.  Need to use quotes for variable names or df\$ syntax
4.  Missing na.rm parameter in mean() function
5.  Need to load ggplot2 package first
6.  Variable scope issues in loops
7.  Need to convert strings to numeric with as.numeric()
8.  NA comparison requires is.na()
9.  Need to add new levels to factors before assigning
10. Use paste() or paste0() for string concatenation

This Rmd file includes:

-   Different difficulty levels
-   Common conceptual mistakes
-   Various R programming concepts
-   Clear descriptions of problems
-   Hidden solutions for class discussion

During class you can: 1. Let students work individually or in pairs 2. Discuss each error and its solution 3. Explain the underlying concepts 4. Share best practices for avoiding these mistakes

Would you like me to adjust any of the examples or add different types of errors?
