---
title: "R Basics - Pre-course Homework"
author: "CRC1678 INF"
date: "2025-07-14"
output: html_notebook
css: "copilot.css"
---

Welcome to R! This exercise will help you understand basic R concepts before our class. Run each code chunk and try to understand the output. Feel free to experiment by adding more chunks.

### About R markdown

This is a R markdown notebook which combines a markdown formatted text and executable R code placed in code chunks. This file has extension `.Rmd` and a new notebook can be created in RStudio by going to File -\> New File -\> R markdown... / R notebook... Try it and you will get a sample notebook which explains how to use markdown and provide you with example code chunks. You can switch between "source" and "visual" modes with buttons on top left - use visual mode to have markdown (especially tables) shown nicely. To run a code chunk, click a small green "play" button on the right of the chunk or use a keyboard shortcut `Ctrl + Shift + Enter` / `Cmd + Shift + Enter`. Practice inserting new chunks and running them.

Full R markdown cheatsheet can be found [here](https://github.com/rstudio/cheatsheets/raw/main/rmarkdown-2.0.pdf).

------------------------------------------------------------------------

### Some useful RStudio Keyboard Shortcuts

Large RStudio cheatsheet can be found at [Posit](https://posit.co/wp-content/uploads/2022/10/rstudio-ide-1.pdf).

You can see all shortcuts by pressing `Alt + Shift + K` /`Option + Shift + K`

**General**

| Action                               | Windows/Linux      | Mac               |
|--------------------------------------|--------------------|-------------------|
| Insert assign operator `<-`          | `Alt + -`          | `Option + -`      |
| Insert tidyverse pipe operator `%>%` | `Ctrl + Shift + M` | `Cmd + Shift + M` |
| Clear line from cursor to beginning  | `Ctrl + U`         | `Cmd + U`         |
| Clear line from cursor to end        | `Ctrl + K`         | `Cmd + K`         |

**Console (lower left window by default)**

| Action                    | Windows/Linux     | Mac              |
|---------------------------|-------------------|------------------|
| Autocompletion            | `Tab`             | `Tab`            |
| Move cursor to console    | `Ctrl + 2`        | `Cmd + 2`        |
| Get previous command      | `arrow-up`        | `arrow-up`       |
| Search command history    | `Ctrl + arrow-up` | `Cmd + arrow-up` |
| Interrupt current command | `Esc`             | `Esc`            |

**Source (upper left window by default)**

| Action | Windows/Linux | Mac |
|----|----|----|
| Move cursor to source | `Ctrl + 1` | `Cmd + 1` |
| Duplicate current line up or down | `Alt + Shift + Up/Down` | `Option + Shift + Up/Down` |
| Delete the current line | `Ctrl + D` | `Cmd + D` |
| Type in multiple lines | `Alt + mouse click + drag` | `Option + mouse click + drag` |
| Comment/Uncomment lines | `Ctrl + Shift + C` | `Cmd + Shift + C` |
| Run current line/selection | `Ctrl + Enter` | `Cmd + Enter` |
| Insert R chunk | `Ctrl + Alt + I` | `Cmd + Option + I` |
| Run current chunk | `Ctrl + Shift + Enter` | `Cmd + Shift + Enter` |
| Save file | `Ctrl + S` | `Cmd + S` |
| Knit document | `Ctrl + Shift + K` | `Cmd + Shift + K` |

------------------------------------------------------------------------

**Notes on typing syntax**

-   When you type R code, remember that R is case sensitive, so `data` is not the same as `Data`.
-   For most cases, it does not matter if you have a space between variable, operator and value, so `2+2` is the same as `2 + 2`. In general it is recommended to use spaces for better code readability.

### Variables assignment

In R (and other languages), variables are used to store values. Basically, they are pointers to an object/location in memory ([Wikipedia](https://en.wikipedia.org/wiki/Variable_(computer_science)). In R, they don't need to be declared and are created at the moment when they are assigned a value.

Things to remember:

-   `<-` is preferred operator for assignment
    -   Use `Alt + -` / `Option + -` shortcut to save typing `<-`.
    -   `=` operator can also be used, but it is discouraged (see this [Extensive R bloggers post](https://www.r-bloggers.com/2018/07/assignment-in-r-slings-and-arrows/) with examples where `=` use for assignment can lead to problems).
    -   Fun fact: `->` assignment also works!
-   variable names:
    -   cannot start with a number
    -   cannot contain `-` (will be evaluated as minus sign)
    -   prefer not to contain spaces, use backticks if you must `` `variable with space` <- 6 ``
    -   should not overwrite built-in functions and variables names

Experiment with assigning variables in the following chunk.

In this chunk we also introduce:

-   `#` character which denotes a comment and can be use to "switch off" lines in your code.
-   `print()` function to output the value of the variable (more on functions below).

```{r}
# Try it
a <- 5 
3 -> b

# Print function prints the values of its variable
print(a)
print(b)

# Reassigning a variable overwrites it
a <- 8
print(a)

# Typing a variable will return its value
b
```

### Functions

Functions in R have a form of `function(argument=value)` and are called as name of the function with parentheses, with respective arguments inside parentheses (here is where `=` operator comes in). There are a number of built-in functions which come with base R (you already saw `print()`).

To obtain help on how to call a function and all possible arguments, type `?function` or `help("function")`. Tip: if you don't remember the exact name, try `apropos("approximate_name")`.

You can use `ls()` function to list your variables and functions - or in RStudio, look at the "Environment" tab (in the top right quadrant by default):

```{r}
ls()
```

To delete an unwanted variable or function, you can use `rm()`

```{r}
rm(a)
ls()
```

You can create your own function in R using the curly brackets syntax:

```{r}
# By including  comment inside our function, we will help user to understand what it does
my_function <- function(a){
  # This function will print the value of the variable given
  print(a)
}
```

And call it:

```{r}
my_function(b)
```

You can see the code of the function by calling its name without brackets:

```{r}
# See the source code of the function
my_function
```

Calling it without an argument leads to an error:

```{r}
my_function()
```

**A note on scope**

You may have noticed that:

-   we have assigned variable `a` and deleted it
-   we have used `a` as an argument inside `my_function` definition and R did not complain that `a` was not found - we have called it on the variable `b` and R did not complain that it wants `a` as input

There are 2 types of variables in R (and other languages): global and local. Global variables can be accessed from anywhere in the code. In our case, `b` is global. Local variables are valid for a part of the code - such as the `a` inside our function. In this case, when the function is called, its argument is assigned to local variable `a`, and this local variable `a` disappears after the function call is finished.

More examples on scope [here](https://www.geeksforgeeks.org/r-language/scope-of-variable-in-r/).

### Data Types in R

Main data types in R are:

-   Numeric types:
    -   integer `1L`, `as.integer(1)`
    -   double `1`, `1.0`, `Inf`
-   Complex `1 + 2i`
-   Character: `"1"`
-   Logical: `TRUE` and `FALSE` (alias `T` and `F`)

You can check a type of variable with `class()` or `str()` (shows a compact representation of object structure).

Note: Special value `NA` (stands for "Not available") represents missing data. By default, `NA` is logical, but it can be converted to other types.

```{r}
class(NA)
```

Additionally, special value `NaN` which stands for "Not a Number". This value is numeric (double) and identifies an undefined number (for example, `0/0`).

You can use operators to perform various operations with these types (tables taken from [w3schools](https://www.w3schools.com/r/r_operators.asp)). To get help on any of the operators, enclose them in backticks (`` ?`%/%` ``)

**Arithmetic**

| Operator | Action                            | Example  | Result |
|----------|-----------------------------------|----------|--------|
| `+`      | Addition                          | `1 + 1`  | `2`    |
| `-`      | Subtraction                       | `3 - 2`  | `1`    |
| `*`      | Multiplication                    | `2 * 3`  | `6`    |
| `/`      | Division                          | `6 / 2`  | `3`    |
| `^`      | Exponent                          | `2 ^ 3`  | `8`    |
| `%%`     | Modulus (Remainder from division) | `7 %% 3` | `1`    |
| `%/%`    | Integer Division                  | `7%/%3`  | `2`    |

```{r}
# Try out arithmetic operators here
9 %/% 5
# Using arithmetic on logical types considers TRUE as 1 and FALSE as 0
TRUE + TRUE
```

**Comparison**

These operators return a logical value.

| Operator | Action                   | Example    | Result  |
|----------|--------------------------|------------|---------|
| `==`     | Equal                    | `3 == 2`   | `FALSE` |
| `!=`     | Not equal                | `"a" != 3` | `TRUE`  |
| `>`      | Greater than             | `5 > 3`    | `TRUE`  |
| `<`      | Less than                | `2 < 3`    | `TRUE`  |
| `>=`     | Greater than or equal to | `1 >= 0`   | `TRUE`  |
| `<=`     | Less than or equal to    | `1 <= 0`   | `FALSE` |

**Logical**

These operators are for logical calculation.

| Operator | Action      | Example        | Result  |
|----------|-------------|----------------|---------|
| `&`,`&&` | Logical AND | `1 & 2`        | `TRUE`  |
| `|`,`||` | Logical OR  | `FALSE | TRUE` | `TRUE`  |
| `!`      | Logical NOT | `! TRUE`       | `FALSE` |

*Note:* yes, single and double logical operators are confusing. The rule of thumb:

-   single forms are "element-wise" and thus are used for vectors (see below).

-   double forms are for `if` statements (see below).

```{r}
# Try out comparison and logical operators here
2 == 3
```

You can use a number of built-in functions to test for specific data type and to convert between data types:

```{r}
# Try out checking if a variable is of specific type - returns logical
is.double(1L)
```

```{r}
# Try converting between types
as.integer("345")
```

### Data structures in R

Single elements can be combined together in a data structure. This table is modified from [Advanced R](http://adv-r.had.co.nz/Data-structures.html) and summarizes them:

| Name | Dimensions | Type | Example command |
|----|----|----|----|
| (atomic) vector | 1D | homogeneous | `num_vector <- c(1,2,3)` |
| list | 1D | heterogeneous | `my_list <- list(1, "foo", TRUE)` |
| matrix | 2D | homogeneous | `m <- matrix(1:4, nrow=2, ncol=2)` |
| array | ND | homogeneous | `a <- array(1:8, dim = c(2,2,2))` |
| data.frame | 2D | heterogeneous | `df <- data.frame(id=1:10, letter=letters[1:10])` |

"Homogeneous" means all elements must be of the same type; heterogeneous allows combining different data types together.

**Vectors and lists**

Vectors and lists are 1-dimensional (they only have length).

```{r}
# Practice vectors

# Create numeric vector
numbers <- c(1, 2, 3, 4, 5)
# ":" operator is a shorthand to create numeric vector of integer range
numbers <- 1:5
print(numbers)
# to get decimal ranges, use the seq() function:
decimals <- seq(from=0.1, to=0.9, by=0.05)
print(decimals)

# Vectorized operations act on each element of the vector:
print(numbers + 2)  # adds 2 to each element
print(numbers * 3)  # multiplies each element by 3

# Find out vector length
print(length(numbers))

# We can assign names to elements of the vector
print(names(numbers)) #returns NULL
names(numbers) <- as.character(1:5) #convert numeric vector to character
print(numbers)
```

*Note:* `NULL` is a specific data type representing "null object". If you create an empty vector with `c()`, it will return `NULL`. See the chapter in [Advanced R](https://adv-r.hadley.nz/vectors-chap.html?q=NULL#null).

**Type coercion**

We said that atomic vectors should contain the same data type. However, if you try to mix data types in a vector, R will not give you an error:

```{r}
my_vector <- c(1,2,"3")
my_vector
```

You see that instead R silently converted integers to characters! This is called type coercion and could lead to some unexpected behavior - see some more examples [here](https://annakrystalli.me/rrresearchACCE20/data-types-structures-and-classes.html#type-coercion).

**Factors**

`factor` in R represents specialized vector to encode categories. Implicitly, it takes a character vector, enumerates all types of words that occur in it, making categories (`levels`) and encodes the vector (internally) as integer with the numbers corresponding to categories.

```{r}
# Practice factors
codons <- as.factor(c("ATG", "ATG", "AAG", "ATG", "AGG", "ATG"))
codons
```

Factors look very similar to character vectors but be aware that they behave differently. Later you will learn `ggplot2` package - it coerces characters to factors for plotting. For modeling, factors are used to encode categorical variables. One useful function that uses factors is `table()` - it counts occurrences for each combination of categories.

```{r}
table(codons)
```

```{r}
# Practice lists here

# Create a list by converting our numeric vector
my_list <- as.list(numbers)
# Lists can be nested (can contain other lists as elements)
nested_list <- list(my_list, 10)
# Name list elements
names(nested_list) <- c("fisrt_list", "addition")
# Length of the list
length(nested_list)
nested_list
```

**Indexing**

R uses 1-based indexing meaning the first element index is 1. Use square brackets to access elements of a vector:

```{r}
# Access elements
print(numbers[1])  # First element
# Putting - in front of the index removes the element:
print(numbers[-2])
# Using : operator returns the range of elements:
print(numbers[2:4])
```

For lists, there is an important distinction: single brackets return a `list` (even if it is one element, it is a list of length 1); use double brackets to return the element itself:

```{r}
# Return a list which is a subset of parent list
print(my_list[1])
print(typeof(my_list[1]))

# Use double bracket to return the element itself
print(my_list[[1]])
print(typeof(my_list[[1]]))

# Set names of the list
# letters is built-in vector or English letters
names(my_list) <- letters[1:5]
# You can use `$` to access an element by name
print(my_list$b)
```

**Matrices, arrays and data.frames**

A lot of data is represented in 2 dimensions (rows and columns), and for this we use matrices and (mostly, in biology) data.frames. For data with more dimensions, we use arrays.

Matrices and arrays only contain 1 data type:

```{r}
m <- matrix(1:12, nrow = 3, ncol = 4)
print(m)
# use dim() to access matrix dimensions
print(dim(m))
# use comma to separate indices while accessing the elements
# [row,col]
print(m[2,3])
# array with 3 dimensions: nrow, ncol, nmat 
# (number of rows, columns and matrices with this number of row/columns)
a <- array(1:24, dim = c(3,4,2))
print(a)
# a matrix is an array (with 2 dimensions)
print(is.array(m))
```

Data.frames can combine multiple types over a row, but column must be of the same type. Essentially, a data.frame can be seen as a list of columns, each of which is a vector of a specific type.

```{r}
# Create a data frame by listing the columns
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 21),
  grade = c(85, 92, 78)
)
print(students)

# Access columns (2 ways):
print(students$name)
print(students[, "age"])

# Access rows
print(students[1, ])  # First row
```

*Note:* with lists and data.frames (but not with vectors, matrices and arrays), you can use `$` notation to access elements by name. In case of data.frames, you can you it to access a column like `data.frame$column_name`. If you get an error `$ operator is invalid for atomic vectors`, that means that you tried to use it on a data structure which is not a list (or a data.frame).

*Note:* a useful vector operator `%in%` allows you to test if some element is present inside a given data structure:

```{r}
10 %in% 1:100
```

**Note on vector recycling**

If you try to create a `data.frame` using vectors of different length, R will not complain, and also not fill in missing values of the shorter vector with `NA`. Instead, it will recycle the shorter vector to fill in missing values, which could lead to unexpected results later. If however the longer vector's length is not a multiple of shorter vector's length, R will return an error:

```{r}
# no warning, x is recycled
print(data.frame(x = 1:2, y = 1:6))
# error
print(data.frame(x = 1:2, y = 1:5))
```

### Statistics and plotting in base R

You will learn `tidyverse` and `ggplot2` packages for tabular data manipulation and plotting, but it is possible to do this with base R functions, too.

### Basic Statistics

Some R built-in functions for basic statistics:

```{r}
data <- c(15, 23, 42, 31, 58)

mean(data)    # Average
median(data)  # Middle value
sd(data)      # Standard deviation
max(data)     # Maximum value
min(data)     # Minimum value
```

**Note on missing values**

As you remember, `NA` stands for missing value and can be of any type, so vectors (and also, lists, arrays and data.frame columns) can contain `NA`. One needs to be aware that `NA` could propagate silenty, because any operation involving `NA` will return `NA`. Use `na.rm=TRUE` argument to ignore missing values:

```{r}
# Example vector with missing values
scores <- c(85, NA, 92, 78, NA)

# Function to for missing values (returns a logical vector)
is.na(scores)

# Calculate mean with missing values
mean(scores)  # Returns NA
mean(scores, na.rm = TRUE)  # Returns a number
```

### Basic Plotting Function

Use `plot()` function to create simple R plots. Most important arguments are `x` , `y` (the data to plot) and `type` (such as dots, lines, histogram) - `?base::plot` is helpful for a quick reference for these and many other parameters.

```{r}
# Create some data
x <- 1:10
y <- x^2

# Make a basic xy point plot
plot(x, y, 
     main = "Simple Plot",
     type = "b",
     xlab = "X axis",
     ylab = "Y axis")
```

### Control flow

Control flow in programming allows you to execute commands based on a condition (`if` and `while` loops) and also to execute the same command on each element of many (`for` loops and `apply` functions).

**Conditions**

Conditions have a form of:

```         
if (condition) {
  perform action
} 
```

If a condition evaluates to `TRUE`, the command under `if` is executed, otherwise it is not, or a different command (under `else`) is executed, if provided.

```{r}
# Simple example of an if-else condition statements
# Try changing the value of x and see what happens
x <- 5
if (x == 5) {
  print("Correct!")
} else {
  print("Wrong!")
}
```

**Loops**

Loops allow you to write a command once instead of repeating it for each element you want to execute it on. A simple loop in R has a form of:

```         
for (element in vector) {
  perform action (element)
}
```

Usually, `for` loops in base R are replaced by a family of `apply` functions. The generic `apply` function has a form of `apply(X, MARGIN, FUNCTION)` where X is the data structure (can also be a matrix or array), MARGIN specifies dimensions to which we want to apply the FUNCTION (for example, 1 means rows, 2 means columns if X is a matrix). The syntax of `apply` is not very intuitive, but we can practice one often used `lapply(X, FUN)` function which performs operation on all elements of a vector or a list:

```{r}
# lapply applies function to a vector/list
# for example, we take a squre root (built-in sqrt() function)
# from the grade from our students dataframe
# lapply always returns a list
lapply(students$grade, sqrt)
```

Some more information on the `apply` family and many examples can be found [here](https://rpubs.com/Richard_h/888425).

**Vectorized operations**

Vectorized functions in R can perform operations on all elements of a vector at the same time, avoiding use of `for` loop and `apply`. Those are very effective, and if possible, it is better to use them instead of loops. Many built-in R functions such as arithmetic operators, `log()` and similar perform vectorized operations automatically. For example, our example with `lapply` is better run like this, returning a simple vector:

```{r}
sqrt(students$grade)
```

**Vectorized conditions**

`ifelse()` is a vectorized form of `if` if you provide a a vector of mixed `TRUE` and `FALSE` as a condition. It has a form of `ifelse(condition_vector, action1, action2)`, where action1 is performed if the condition in `TRUE` and action2 if the condition is `FALSE`. For example, we can create a vector based on student's grade:

```{r}
ifelse(students$grade>80, "excellent", "poor")
```

**Pipe**

Pipe is an operator to direct a variable or a result of a statement into a next statement, allowing to chain several statements after each other. The pipe of the form `%>%` comes with the `tidyverse` package, however, base R has a built-in pipe of the form `|>` (since R 4.0). Mainly the pipe serves to make your code more readable and allows to better break it over several lines:

```{r}
# instead of nesting functions like log(sum(sqrt(numbers))), you can write:
numbers |>
  sqrt() |>
  sum() |>
  log()
```
