---
title: "R Basics - Pre-course Homework"
author: "CRC1678 INF"
date: "2025-07-14"
output: html_document
css: "copilot.css"
---

Welcome to R! This exercise will help you understand basic R concepts before our class. Run each code chunk and try to understand the output. Feel free to experiment by adding more chunks.

### About R markdown

This is a R markdown notebook which combines a markdown formatted text and executable R code placed in code chunks. This file has extension `.Rmd` and a new notebook can be created in RStudio by going to File -\> New File -\> R markdown... Try it and you will get a sample notebook which explains how to use markdown and provide you with example code chunks. To run a code chunk, click a small green "play" button on the right of the chunk or use a keyboard shortcut `Ctrl + Shift + Enter` / `Cmd + Shift + Enter`. Practice inserting new chunks and running them.

Full R markdown cheatsheet can be found [here](https://github.com/rstudio/cheatsheets/raw/main/rmarkdown-2.0.pdf).

------------------------------------------------------------------------

### Some useful RStudio Keyboard Shortcuts

Large RStudio cheatsheet can be found at [Posit](https://posit.co/wp-content/uploads/2022/10/rstudio-ide-1.pdf).

You can see all shortcuts by pressing `Alt + Shift + K` /`Option + Shift + K`

**General**

| Action                               | Windows/Linux      | Mac               |
|--------------------------------------|--------------------|-------------------|
| Insert assign operator `<-`          | `Alt + -`          | `Option + -`      |
| Insert tidyverse pipe operator `%>%` | `Ctrl + Shift + M` | `Cmd + Shift + M` |
| Clear line from cursor to beginning  | `Ctrl + U`         | `Cmd + U`         |
| Clear line from cursor to end        | `Ctrl + K`         | `Cmd + K`         |

**Console (lower left window by default)**

| Action                    | Windows/Linux     | Mac              |
|---------------------------|-------------------|------------------|
| Autocompletion            | `Tab`             | `Tab`            |
| Move cursor to console    | `Ctrl + 2`        | `Cmd + 2`        |
| Get previous command      | `arrow-up`        | `arrow-up`       |
| Search command history    | `Ctrl + arrow-up` | `Cmd + arrow-up` |
| Interrupt current command | `Esc`             | `Esc`            |

**Source (upper left window by default)**

| Action | Windows/Linux | Mac |
|----|----|----|
| Move cursor to source | `Ctrl + 1` | `Cmd + 1` |
| Duplicate current line up or down | `Alt + Shift + Up/Down` | `Option + Shift + Up/Down` |
| Delete the current line | `Ctrl + D` | `Cmd + D` |
| Type in multiple lines | `Alt + mouse click + drag` | `Option + mouse click + drag` |
| Comment/Uncomment lines | `Ctrl + Shift + C` | `Cmd + Shift + C` |
| Run current line/selection | `Ctrl + Enter` | `Cmd + Enter` |
| Insert R chunk | `Ctrl + Alt + I` | `Cmd + Option + I` |
| Run current chunk | `Ctrl + Shift + Enter` | `Cmd + Shift + Enter` |
| Save file | `Ctrl + S` | `Cmd + S` |
| Knit document | `Ctrl + Shift + K` | `Cmd + Shift + K` |

------------------------------------------------------------------------

**Notes on typing syntax**

-   When you type R code, remember that R is case sensitive, so `data` is not the same as `Data`.
-   For most cases, it does not matter if you have a space between variable, operator and value, so `2+2` is the same as `2 + 2`. In general it is recommended to use spaces for better code readability.

### Variables assignment

In R (and other languages), variables are used to store values. Basically, they are pointers to an object/location in memory ([Wikipedia](https://en.wikipedia.org/wiki/Variable_(computer_science)). In R, they don't need to be declared and are created at the moment when they are assigned a value.

Things to remember:

-   `<-` is preferred operator for assignment
    -   Use `Alt + -` / `Option + -` shortcut to save typing `<-`.
    -   `=` operator can also be used, but it is discouraged (see this [Extensive R bloggers post](https://www.r-bloggers.com/2018/07/assignment-in-r-slings-and-arrows/) with examples where `=` use for assignment can lead to problems).
    -   Fun fact: `->` assignment also works!
-   variable names:
    -   cannot start with a number
    -   cannot contain `-` (will be evaluated as minus sign)
    -   prefer not to contain spaces, use backticks if you must `` `variable with space` <- 6 ``
    -   should not overwrite built-in functions and variables names

Experiment with assigning variables in the following chunk.

In this chunk we also introduce:

-   `#` character which denotes a comment and can be use to "switch off" lines in your code.
-   `print()` function to output the value of the variable (more on functions below).

```{r}
# Try it
a <- 5 
3 -> b

# Print function prints the values of its variable
print(a)
print(b)

# Reassigning a variable overwrites it
a <- 8
print(a)

# Typing a variable will return its value
b
```

### Functions

Functions in R have a form of `function(argument=value)` and are called as name of the function with parentheses, with respective arguments inside parentheses (here is where `=` operator comes in). There are a number of built-in functions which come with base R (you already saw `print()`).

To obtain help on how to call a function and all possible arguments, type `?function` or `help("function")`. Tip: if you don't remember the exact name, try `apropos("approximate_name")`.

You can use `ls()` function to list your variables and functions - or in RStudio, look at the "Environment" tab (in the top right quadrant by default):

```{r}
ls()
```

To delete an unwanted variable or function, you can use `rm()`

```{r}
rm(a)
ls()
```

You can create your own function in R using the curly brackets syntax:

```{r}
# By including  comment inside our function, we will help user to understand what it does
my_function <- function(a){
  # This function will print the value of the variable given
  print(a)
}
```

And call it:

```{r}
my_function(b)
```

You can see the code of the function by calling its name without brackets:

```{r}
# See the source code of the function
my_function
```

Calling it without an argument leads to an error:

```{r}
my_function()
```

**A note on scope**

You may have noticed that:

-   we have assigned variable `a` and deleted it
-   we have used `a` as an argument inside `my_function` definition and R did not complain that `a` was not found - we have called it on the variable `b` and R did not complain that it wants `a` as input

There are 2 types of variables in R (and other languages): global and local. Global variables can be accessed from anywhere in the code. In our case, `b` is global. Local variables are valid for a part of the code - such as the `a` inside our function. In this case, when the function is called, its argument is assigned to local variable `a`, and this local variable `a` disappears after the function call is finished.

More examples on scope [here](https://www.geeksforgeeks.org/r-language/scope-of-variable-in-r/).

### Data Types in R

Main data types in R are:

-   Numeric types:
    -   integer `1L`, `as.integer(1)`
    -   double `1`, `1.0`, `Inf`
-   Complex `1 + 2i`
-   Character: `"1"`
-   Logical: `TRUE` and `FALSE` (alias `T` and `F`)

You can check a type of variable with `class()` or `str()` (shows a compact representation of object structure).

Note: Special value `NA` (stands for "Not available") represents missing data. By default, `NA` is logical, but it can be converted to other types.

```{r}
class(NA)
```

Additionally, special value `NaN` which stands for "Not a Number". This value is numeric (double) and identifies an undefined number (for example, `0/0`).

You can use operators to perform various operations with these types (tables taken from [here](https://www.w3schools.com/r/r_operators.asp):

**Arithmetic**

| Operator | Action                            | Example  | Result |
|----------|-----------------------------------|----------|--------|
| `+`      | Addition                          | `1 + 1`  | `2`    |
| `-`      | Subtraction                       | `3 - 2`  | `1`    |
| `*`      | Multiplication                    | `2 * 3`  | `6`    |
| `/`      | Division                          | `6 / 2`  | `3`    |
| `^`      | Exponent                          | `2 ^ 3`  | `8`    |
| `%%`     | Modulus (Remainder from division) | `7 %% 3` | `1`    |
| `%/%`    | Integer Division                  | `7%/%3`  | `2`    |

```{r}
# Try out arithmetic operators here
9 %/% 5
```

**Logical**

| Operator | Action | Example | Result |
|----------|--------|---------|--------|

| `==` \| Equal \| `3 == 2` \| `FALSE` \|
| `!=` \| Not equal \| `"a" != 3` \| `TRUE` \|
| `>` \| Greater than \| `5 > 3` \| `TRUE` \|
| `<` \| Less than \| `2 < 3` \| `TRUE` \|
| `>=` \| Greater than or equal to \| `1 >= 0` \| `TRUE` \|
| `<=` \| Less than or equal to \| `1 <= 0` \| `FALSE` \|

```{r}
# Try out logical operators here
2 == 3
```

You can use a number of built-in functions to test for specific data type and to convert between data types:

```{r}

```

## 3. Vectors

Vectors are collections of elements of the same type:

-   `NULL` <https://adv-r.hadley.nz/vectors-chap.html?q=NULL#null> (e.g. empty `c()`)

```{r}
# Create numeric vector
numbers <- c(1, 2, 3, 4, 5)
print(numbers)

# same can be done with 
x <- 1:5
# Create character vector
fruits <- c("apple", "banana", "orange")
print(fruits)

# Vector operations
numbers + 2  # adds 2 to each element
numbers * 3  # multiplies each element by 3

# Access elements
print(numbers[1])  # First element
print(fruits[2])   # Second element
```

Indexing: R uses 1 -based indexing

vector operators: :, %in%, %\*%(matmul)

Create your own vectors and try operations:

```{r}
# Write your code here

```

## 5. Data Frames

Data frames are like spreadsheets in R:

```{r}
# Create a simple data frame
students <- data.frame(
  name = c("Alice", "Bob", "Charlie"),
  age = c(20, 22, 21),
  grade = c(85, 92, 78)
)

# Look at the data
print(students)

# Access columns
print(students$name)
print(students[, "age"])

# Access rows
print(students[1, ])  # First row
```

Create your own data frame about any topic:

```{r}
# Write your code here

```

### Missing Values

R uses `NA` to represent missing values:

```{r}
# Vector with missing values
scores <- c(85, NA, 92, 78, NA)

# Check for missing values
is.na(scores)

# Calculate mean with missing values
mean(scores)  # Returns NA
mean(scores, na.rm = TRUE)  # Ignores NA
```

Try working with missing values:

```{r}
# Write your code here

```

### Basic Statistics

R has built-in functions for basic statistics:

```{r}
data <- c(15, 23, 42, 31, 58)

mean(data)    # Average
median(data)  # Middle value
sum(data)     # Sum of all values
length(data)  # Number of elements
max(data)     # Maximum value
min(data)     # Minimum value
```

Try these functions with your own numbers:

```{r}
# Write your code here

```

### Plotting

Let's create a simple plot:

```{r}
# Create some data
x <- 1:10
y <- x^2

# Make a basic plot
plot(x, y, 
     main = "Simple Plot",
     xlab = "X axis",
     ylab = "Y axis")
```

Try creating your own plot:

```{r}
# Write your code here

```

### Control flow

if an for loops

vectorized operations

internal pipe (apply to self made functions)

```{r}

```
